---
title: "Homework 10"
author: "Jina Moon"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: no
---

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!    

## Background
In the `data` folder, you will find data about shark incidents in California between 1950-2022. The [data](https://catalog.data.gov/dataset/shark-incident-database-california-56167) are from: State of California- Shark Incident Database.   

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Load the data
Run the following code chunk to import the data. In this chunk, I am removing NA's.
```{r message=FALSE, warning=FALSE}
sharks <- 
  read_csv("data/SharkIncidents_1950_2022_220302.csv", 
           na=c("NA", "N/A", "Unknown", "NOT COUNTED", "", " ")) %>% 
  clean_names()
```

```{r}
glimpse(sharks)
```

## Questions
1. Are there any "hotspots" for shark incidents in California? Show the top five counties with the most incidents.
```{r}
sharks %>% 
  select(county) %>%  
  group_by(county) %>% 
  summarize(incident_total = n()) %>% 
  slice_max(incident_total, n = 5)
```

2.  Make a plot that shows the total number of incidents per county.
```{r}
#basically just a barplot 
sharks %>% 
  select(county) %>%  
  group_by(county) %>% 
  summarize(incident_total = n()) %>% 
  ggplot (aes(x = county, y = incident_total)) +
  geom_col(aes(fill = county)) + 
  labs(title = "Total number of incidents per county", x = "county", y ="total number of incidents") + 
  coord_flip()
```

3. Are there months of the year when incidents are more likely to occur? Make a plot that shows the total number of incidents by month. Which month has the highest number of incidents?  
```{r}
#hmm probably bar plot 
sharks %>% 
  mutate(month = as.factor(month)) %>% #convert to factors since barplot 
  select(month) %>% 
  group_by(month) %>% 
  summarize(incident_total = n()) %>% 
  ggplot (mapping=aes(x=month, y=incident_total)) + 
  geom_col(aes(fill = month)) + 
  labs(title = "Number of incidients by month", x = "month", y = "total # of incidents") + 
  theme(plot.title = element_text(size=rel(1.5), hjust = .5)) 
```

Yes there are months of the year where incidents are more likely to occur (mainly summer months, or ones near it). October has the highest number of incidents. 


4. Is there a time of day when most attacks occur? Make a plot that shows the total number of incidents by time of day. Which time of day has the highest number of incidents?  
```{r}
#we'll just remove the non numbers
#prob also barplot since has to do with count 

sharks %>% 
  separate(time, into= c("hour", "min"), sep=":") %>% 
  mutate(hour = as.numeric(hour)) %>% #gets rid of non-numbers
  select(hour) %>%
  mutate(hour = factor(hour, levels = 23:0)) %>% #specified levels that I want using factor 
  group_by(hour, .drop = F) %>% #added .drop = F to make it all values show up
  summarize (incident_total = n()) %>% 
  ggplot (mapping=aes(x=hour, y=incident_total)) + 
  geom_col(aes(fill = hour)) + 
  labs(title = "Number of incidients by time of day (hour)", x = "time of day (hour)", y = "total # of incidents") + 
  theme(plot.title = element_text(size=rel(1.5), hjust = .5)) + 
  coord_flip()
```

The most attacks seem to occur between 11 am and 12 pm. 
For the plot I chose to ignore the min to make it more... visually readable. 

5. How do the number and types of injuries compare by county? Make a table that shows the number of injury types by county. Which county has the highest number incidents? (hint: your table should be in wide format)
```{r}
#table that shows # injury types by county 
sharks_wide <- sharks %>% 
  group_by(county, injury) %>% 
  summarize(incident_total_type = n()) %>% 
  pivot_wider(names_from= "injury",
              values_from="incident_total_type") %>% 
  mutate(across(where(is.numeric), ~ifelse(is.na(.), 0, .)))  %>%  #replaces na with 0 which works in this context
  mutate(total_incidents = minor + none + major + `minor*` + `none*` + fatal + `major*`) %>% 
  arrange(desc(total_incidents))
sharks_wide
```

San Diego... perhaps unsurprisingly... has the highest number incidents. 


6. Use the table from #5 to make a plot that shows the total number of incidents by county.
```{r}
sharks_wide %>% 
  ggplot(aes(x = county, y = total_incidents)) +
  geom_col(aes(fill = county)) +
  coord_flip() + 
  labs(title = "Total # of Incidents by County", x = "County", y = "Total Incidents")
```

7. In the data, `mode` refers to a type of activity. Which activity is associated with the highest number of incidents?
```{r}
sharks %>% 
  group_by(mode) %>% 
  summarize(activity_total = n()) %>% 
  arrange(desc(activity_total))
```

Surfing/Boarding is associated with the highest number of incidents. 


8. Make a plot that compares the number of incidents by activity.
```{r}
#bar plot but col since specific y
sharks %>% 
  group_by(mode) %>% 
  summarize(activity_total = n()) %>% 
  ggplot(aes(x = mode, y = activity_total)) +
  geom_col(aes(fill = mode)) +
  coord_flip() +
  labs(title = "Number of Incidents by Activity", x = "Activity", y = "Total Incidents")

```

9. Which shark species is involved in the highest number of incidents?  
```{r}
sharks %>% 
  mutate(species = tolower(species)) %>% #Blue is a weirdo 
  group_by(species) %>% 
  summarize(total_incident = n()) %>% 
  arrange(desc(total_incident))
```

The Great white shark is involved in the highest number of incidents. 

10. Are all incidents involving Great White's fatal? Make a plot that shows the number and types of incidents for Great White's only.  
```{r}
sharks %>% 
  filter(species == "White") %>%  
  group_by(injury) %>% 
  summarize(incident_total_white = n()) %>% 
  ggplot(aes(x = injury, y = incident_total_white)) +
  geom_col(aes(fill = injury)) +
  labs(title = "Incidents involving Great White Shark", x = "Injury", y = "Incident Total")
```

No, not all incidents involving Great White sharks are fatal, and most of it tend to not be fatal. 


## Knit and Upload
Please knit your work as an .html file and upload to Canvas. Homework is due before the start of the next lab. No late work is accepted. Make sure to use the formatting conventions of RMarkdown to make your report neat and clean! 