---
title: "BIS15L Notes"
author: "Jina Moon"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
---

##Computer Hygeine

Sometimes I suppose it is good to keep my desktop, tabs, and physical laptop clean– 

##GitHub

https://github.com/MoonJ05/BIS15L_W26_jmoon.git ← link to my repository 

##Keyboard short cuts (mac)

* Option + “-” gets symbol (<-) 
* Command + option + i creates a code chunk 
* Command + return runs
* command + shift + 'm' gets %>% 

##Libraries to load
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")  #for cleaning
library("ggthemes") #for themes
library("naniar") #for NAs
library("RColorBrewer")
library("paletteer") #for color palettes 
library("lubridate") #this will help us manage dates
library(ggmap) #for maps
library(leaflet) #for leaflet 
library("shiny") #shiny app
library(shinydashboard)
library(shinythemes)


#Stamen API key
register_stadiamaps("bf91d81b-2ad3-45b1-a217-21bb0a13f297", write = FALSE)
#needed for maps 
``` 
  
##Basic commands

```{r}
getwd()    #checks current directory 
setwd("") 	#changes your working directory to "" 
write.csv(data_name, "datafile.csv", row.names=FALSE)	#row.names avoid row numbers from printing out. Saves data frame data_name as .csv file 
Data_name <- read.csv("datafile.csv")	 #import .csv file 
#use read_csv
y <- 3 		#assigns new value to object
?round 		#learn how to use function (ex. round) 
X <- c(1, 2, 3)		#vector… ! can store multiple values (concatenate) 
# ex. char_vector <- c("Monday", "Tuesday")
#     Sequence <- c(1:100) 

X[2] 		#outputs a 2… vector starts at index 1
X <=10		#outputs T/F for all values in vector X <= to 10 
x[x<=10]	#using ‘[ ]’ outputs values not bool of entire vector
# <, >, ==, <=, >= 
Data <- data		#reads data into R to visualize 
```

##Tidyverse

Collection of R packages designed for data science. 
* Load with library(“tidyverse”)
```{r}
#Creating data frame with tibble: 
#First create vectors you want in the data frame in same orders
Name_data <- tibble(vector1, vector2, vector3) 
# ex. birds <- tibble (sex, length, weight) 
```


##Dplyr

Used to transform data frames by extracting, rearranging, and summarizing data. 

```{r}
#use 'df(dataframe) <- ' to save changes concretely 
select(data_name, var_1, var_2)	  #pull out columns of interest from the dataframe.. Order matters 
select(data_name, start_col_name:end_col_name) 		#pulls all the between too 
select(data_name, !var_1)		#pulls all except specified
select(data_name, !c(var_1, var_2))	#applies to all in vector
select(data_name, contains ("mm")) 	#pulls if has mm
    starts_with("a")
    ends_with("z")
    where(is.numeric()) 		#based on class of data
rename(data_name, new_name = old_name) 
select(data_name, var1, var2New = var2) #does the same thing as rename above.

relocate(data_frame, var4) 		#relocates variable to first column
filter(data_name, var== "specific_cat")	#extracts data that meet criteria within row 
var_1 >= 400
# >, >=, <, <=, !=, ==
Var_1 %in% c(5250, 5500) 		#restricts to these two values (exactly)
between(var, value_min, value_max) 
near(var, 5400, tol=100)		#splits into 50 left, 50 right! 5350 to 5450 
#Conditional operators if multiple conditions
#   & 	checks for both …. Separating with “,” defaults to &
#   |	 “or” 
```

##Pipies

Feed the output from one function into the input of another function 

```{r}
#ex.  
Data %>%
  select(data, var_1) %>%
  filter(var_1 == "specific") %>%
  arrange(var_1) 
#general rule select goes first. Arrange defaults to ascending ,but can do desc()

slice_max(var_1, n = 1) 	#outputs n max
slice_head(var_1, n = 1) 
slice_min()
slice_tail() 
```

##Data Frames

```{r}
names(data) 		#gives you names of vector (column) 
dim(data)		#gives dimensions
str(data)
summary(data) 	#summarizes data frame
glimpse(data)		#gives basic
nrow(data)		#gives number of rows
ncol(data)		#gives number column 
head(data, n)		#prints first n  rows of data frame
tail(data, n)		#prints the last n rows of data frame
Data[row, column]		#can do data[1,] for first row, data[,3] for third column
mean(hbirds$length)		#use $ to access column (var) in data frame 

#To add new data to the data frame… 
Data <- tibble(var1, var2,var3,...)	#recreate whole data
#Can also do …. 
data$nameNewVar <- NewVar
Add_column(data, newVarVector)
#if you do / + * - to a vector using a different vector, it automatically creates a new vector with values 

distinct() 	#outputs all the unique observations in rows. If there is no duplicate of any mix of data, all should print. 
Dataframe %>% distinct(var1)	#looks at var1 and unique observations of that variable, in the data frame.
distinct(var1, var2, .keep_all = T) #looks at both… last part makes sure other part of row printed too

#add new column 
mutate(new_varName = old / 1000) #create new columns from existing ones in the data frame. Added to end by default. 
mutate(across(ends_with("mm"), ~./10)) #applies to all columns ending with ‘mm’
~./10 #lambda function (formula)... ‘.’ current column… 
```

##Data Rules / Cleaning Data

```{r}
#check lab 8 fish clean for more! 
#Use lowercase (everything)
#no spaces!
#seperate data into valriables (row), values, and columns (observations)
rename() #to change column names or using select() 
Data_name <- data_name %>% clean_names()  #Cleans name and fixes everything! All lower case and spaces to underscores. 
#For observations itself: 
Dataframe %>% mutate(across(c("var1", "var2"), tolower)) #to lower case… but numeric becomes char if included… 
mutate(across(c(years_of_sample_collection, elephant_id, sex), as.factor)) #when changing type.. very similar 
Datafame %>% mutate(across(where(is.character), tolower)) #safer

#useful code: 
elephants <- read_csv("data/elephants.csv") %>% 
  clean_names() %>% 
  mutate(across(where(is.character), as.factor))

#Using ifelse… 

#ifelse(logical statement, true, false) … 
#ex. 
ifesle(newborn == -999.00, NA, newborn)
mutate(newborn_new=ifelse(newborn == 999, NA, newborn))
#For ALL
mutate(across(c(a,b,c) ~ifelse(.== -999.00, NA, .))) 
#‘.’ is current. ~ to start formula 
```

##Types of Data

```{r}
Numeric (42) #Dbl means double precision floating point 
Integer (2L) #L denotes integer (whole number) 
Character ("universe")
Logical (T/F) 

#Factors: categorical variables that have a fixed number of possible values called levels
#Discrete: quantitative data w/ only integers
#Continuous: quantitative data with any numerical value 
#Categorical: qualitative data with limited number of values

class(object)		#gives you what type of data you are working with 
is.integer(object)	#checks T/F if specific type
as.integer(object)	#converts to integer
hotsprings$scientist <- as.factor (hotsprings$scientist)	#converts to factor 
#to do this for the whole row...
mutate(across(where(is.character), as.factor))

levels(data$variable)	#checks for level
table(data$variable)	#gives level and count
# NA is filled in for missing … is.na(object) checks boolean 
anyNA(object)		#checks of any NA in vector 
mean(vector, na.rm = T) 
```

##Graphs

```{r}
#Basic idea: plot = data + geom_ + aesthetics. 
#Use scale_x_log10() to scale x-axis
#Use scale_y_log10() to scale y-axis 

#Box plot (categorical vs. continuous) !!
ggplot (data=data_name, 
	mapping=aes(x=x_var_name, y=y_var_name)) + 
	geom_boxplot(mapping=aes(fill = categorical_var_name)) +
	labs (title = "x (unit) vs. y", x= "x_var", y = "y_var") 
# *remember. Box plot = median and IQR… NOT mean. 

#Scatter plots (continuous vs. continuous) 
ggplot (data = data_name, 
        mapping=aes(x=x_var_name, y=y_var_name)) + 
  geom_point (na.rm = T, mapping = aes (color = var_name, shape=var_name)) + 
  geom_smooth (method = lm, se = T) + 
  labs (title = "title", x = "name", y = "name", color = "name")
#… color can make data points a different color like species differentiate (adding within the initial one makes multiple regression lines, doing it with geom_point makes color only applied to points. na.rm=T removes NA warning, geom_smooth adds regression line, and se=T adds standard error
#x is IV, y is DV 

#Barplots (counts # observation for one categorical) 
ggplot(data = data_name, 
       mapping = aes (x = x_var)) + 
  geom_bar (mapping = aes(fill = x_var_name)) + 
  labs (title = "title", x = "x", y = "count")
#use fill instead of color to fill inside bars. Fill can be a different categorical variable. 
#mapping= aes() 		#anytime you want to add color or shape 

#Barplot when we have specific y
msleep %>% 
  count(vore) %>% 
  ggplot(aes(x=vore, y=n)) + #count(vore) gives n 
  geom_col(mapping = aes (fill = x_var_name)) #use this when we have an x and a y... geom_bar() is used when no y. 

```

##Statistical Summaries

```{r}
#when you want summaries in one line. 
#filter when you want a specific value from variable
#select when you want to narrow variables you are looking at

msleep %>% 
  filter(bodywt>150) %>% 
  summarize(mean_bodywt_lg=mean(bodywt))

msleep %>% 
  filter(bodywt>150) %>% 
  summarize(mean_bodywt_lg=mean(bodywt), 
            min_bodywt_lg=min(bodywt), 
            max_bodywt_lg=max(bodywt),
            sd_bodywt_log=sd(bodywt),
            total=n())

msleep %>% 
  select(where(is.numeric)) %>%
  summarize_all(mean, na.rm = T)
#what they are trying to move people toward 
msleep %>% 
  summarize(across(where(is.numeric),
                   ~mean(.x, na.rm=T))) #.x means for every cell

msleep %>% 
  filter(bodywt>100) %>% #check to make sure each line is working before moving on 
  distinct() #gives list of distinct

#distinct.... count 
msleep %>% 
  filter(bodywt>100) %>% 
  summarize(n_genera=n_distinct(genus))

#total obsevations
msleep %>% 
  filter(bodywt>100) %>% 
  summarize(total=n())

msleep %>% 
  filter(bodywt > 100 ) %>% 
  distinct(genus)
#as vector
msleep %>% 
  filter(bodywt>100) %>%
  distinct() %>% #if you also put distinct (genus) it only outputs distinct 
  pull(genus) #gives vector for pull. 
#count observations 
msleep %>% 
  count(vore)

msleep %>% 
  count(vore, order)
msleep %>%  
  tabyl(vore) #gives perent on top of n 

msleep %>% 
  count(vore) %>% 
  ggplot(aes(x=vore, y=n)) + #count(vore) gives n 
  geom_col() #use this when we have an x and a y... geom_bar() is used when no y. 
```

##Group_by

```{r}
#utilized to more easily calculate mean, sd, etc. for different groups
#can group by more than one variable. 
elephants %>% 
  group_by(years_of_sample_collection, sex) %>% 
  summarize(mean_age = mean(estimated_age_years, na.rm = T), 
            meadian = median(estimated_age_years, na.rm = T), 
            standard_dev = sd(estimated_age_years, na.rm = T), 
            n = n()) %>% 
  arrange(sex)
#if need to filter, always filter first. 
elephants %>% 
  group_by(years_of_sample_collection, sex) %>% 
  summarize(mean_estimated_age_years = mean(estimated_age_years, na.rm = T), 
n=n(), 
.groups='keep') #makes error message go away 



#if you run into one that gives weird numbers... you can view them this way: 
msleep %>% 
  filter(order == "Cetacea") %>% 
  select(order, genus, brainwt) 

#interesting one for reference 
fisheries_clean %>% 
  filter(period=="2023") %>%
  group_by(common_name) %>% 
  summarize(sum_catch = sum(catch, na.rm = T)) %>% 
  arrange(desc(sum_catch)) %>% 
  head(sum_catch, n =10)
```

##Learning more about data 

```{r}
#to figure out num different
summarize(n_distinct=n_distinct(column_name)) # you can summarize more than one, but comma
#other applicaitons fo summarize
summarize(sum = sum(catch, na.rm = T))

```

##Pivoting Data
```{r}
#up and down = variables. 
#left to right are the observations
#individually they are values 

#pivot longer: when column names represent variables. 

#if the column (aka variables), need to be part of observations use pivot longer e. 
heartrate_long <- heartrate %>% 
  pivot_longer(-patient, #what stays fixed
               names_to="drug", #puts the column names under this name instead
               values_to = "heartrate",
               values_drop_na = T) #the values are redefined more precisely 
#in pivot longer there are different ways to pivot. in the first section you can...
wk1:wk76 #specify range you want to make longer
-c(x,y,z) #specify columns that you want to stay fixed
#identity columns by prefix, remove prefix  
billboard %>% 
  pivot_longer(
    cols=starts_with("wk"),
    names_to = "week",
    names_prefix = "wk", #gets rid of the prefix 
    values_to="rank",
    values_drop_na = T
  )
#when more than one variable in a column name... 
data %>% 
  pivot_longer(
    letter_a:letter_b, #for these columns
    names_to = c("letter", "alphabet") #new variable names (what they will split into)
    names_sep = "-" #what the names are seperated by 
    values_to = "mRna" #aka what the values are actually a stand in for (bbp, etc.)
  )
#to make it more tidy, and get rid of prefix... 
qpcr_tidy %>% 
  separate(experiment, into = c("experiment", "e_number"), sep=3) %>% #sep = 3, is what index it seperates at.... exp1... at 3 so at p. 
  separate(replicate, into=c("rep", "r_number"), sep = 3) %>% 
  select(gene, experiment=e_number, replicate = r_number, 'mRNA expression')

#seperate()
  seperate (patient, into=c("patient", "gender"), sep = "_") #column, new column, separator
#we can also do sep = -1 (split at the last character)
#further clean patient_f by...
  mutate(patient=str_remove(patient, "_")) #mutates
#unite is the opposite of seperate()...
  unite(patient_gender, "patient", "gender", sep="_")
  
#pivot wider... use when variable names are nder values. 
  pivot_wider(names_from = "key", #new column names 
              values_from= "value") 
```

## REVIEW of plots 2.1 
```{r}
#geom_bar. when you only have one variable

ggplot(data = life_history, 
       mapping = aes (x = order)) + 
  geom_bar (mapping = aes(fill = order)) + 
  labs (title = "order count", x = "order", y = "count") +
  coord_flip() #add so that you can actually read the order names 

#geom_col for when you have x and y variable 

life_history %>% 
  count(order) %>% 
  ggplot(aes(x=order, y =n)) + 
  geom_col(aes(fill = order)) +
  coord_flip()

#bar plot using summarize 

life_history %>% 
  group_by(order) %>% 
  summarize(avg_mass = mean(mass,  na.rm = T)) %>% #adds another variable for order group
  arrange(desc(avg_mass)) %>% 
  ggplot(aes(x = order, y = avg_mass)) + #ggplot(aes(x=reorder(order, avg_mass), y = avg_mass)) reorders it so x-axis label is more readable 
  geom_col(aes(fill = order)) + 
  coord_flip() +
  scale_y_log10()
options(scipen=999) #cancels scientific notation for the session

#another way of scaling
ggplot (data= life_history, 
	mapping=aes(x= order, y=log10(mass)))

#scatter plot (two continuous)
life_history %>% 
ggplot (mapping=aes(x=mass, y=litter_size)) + 
  geom_point (na.rm = T) + 
  scale_x_log10()+
  geom_smooth (method = lm, se = T, na.rm = T) + 
  labs( title="Mass vs. litter size", 
        x="Mass(g)", 
        y="Litter size") + #y = NULL means no label at all 
  theme(plot.title = element_text(size=rel(1.5), hjust = .5)) #making it relative to others by 1.5 times bigger.. also adjust horizontal placement to .5 (middle)


#aesthetics of plots
geom_point (size = 3, color = "black", alpha=0.8, shape = 24, fill = "blue")transparency #sizing and color, etc. alpha controls transparency (shape is 0 to 24) 

colors() #shows all the colors

homerange %>% 
  count(taxon, sort = T) #sorts! whoaaa. rearrange from largest to smallest

#stacked bar plot (abomination )
homerange %>% 
  ggplot(aes(x = taxon, fill=trophic_guild)) + #abomination of the century 
  geom_bar()
#better = dodge 
homerange %>% 
  ggplot(aes(x = taxon, fill=trophic_guild)) + #abomination of the century 
  geom_bar(position = "dodge")+ #always better to hae side by side
#to orient vertically: 
    theme(axis.text.x = element_text(angle = 45, hjust=0.85)) #whoa 

#can scale to percentage too. 
homerange %>% 
  ggplot(aes(x=taxon, fill = trophic_guild)) + 
  geom_bar(position=position_fill()) + 
  scale_y_continuous(labels=scales::percent)+
  coord_flip()
# GET RID OF NA
na.omit()

#group. 
#similar to fill but without color. Groups data for plotting 
homerange %>% 
  ggplot(aes(x=class, y =log10_mass, group= taxon)) + #can be useful, but fill does the same thing but better mot of the times... 
  geom_boxplot()


#some examples of applicaiton 
#number incidents per month. 
sharks %>% 
  mutate(month = as.factor(month)) %>% #convert to factors since barplot 
  select(month) %>% 
  group_by(month) %>% 
  summarize(incident_total = n()) %>% 
  ggplot (mapping=aes(x=month, y=incident_total)) + 
  geom_col(aes(fill = month)) + 
  labs(title = "Number of incidients by month", x = "month", y = "total # of incidents") + 
  theme(plot.title = element_text(size=rel(1.5), hjust = .5)) 

#time of day 
sharks %>% 
  separate(time, into= c("hour", "min"), sep=":") %>% 
  mutate(hour = as.numeric(hour)) %>% #gets rid of non-numbers
  select(hour) %>%
  mutate(hour = factor(hour, levels = 23:0)) %>% #specified levels that I want using factor 
  group_by(hour, .drop = F) %>% #added .drop = F to make it all values show up
  summarize (incident_total = n()) %>% 
  ggplot (mapping=aes(x=hour, y=incident_total)) + 
  geom_col(aes(fill = hour)) + 
  labs(title = "Number of incidients by time of day (hour)", x = "time of day (hour)", y = "total # of incidents") + 
  theme(plot.title = element_text(size=rel(1.5), hjust = .5)) + 
  coord_flip()

#table that shows # injury types by county 
sharks_wide <- sharks %>% 
  group_by(county, injury) %>% 
  summarize(incident_total_type = n()) %>% 
  pivot_wider(names_from= "injury",
              values_from="incident_total_type") %>% 
  mutate(across(where(is.numeric), ~ifelse(is.na(.), 0, .)))  %>%  #replaces na with 0 which works in this context
  mutate(total_incidents = minor + none + major + `minor*` + `none*` + fatal + `major*`) %>% 
  arrange(desc(total_incidents))
sharks_wide
```

## Line plots, histograms, and density plots

```{r}
#for a scatter plot with line 
gap %>% 
  filter((year>=2000 & year <= 2025) & country == "USA") %>% 
  ggplot(aes(x = year, y = gdp)) +
  geom_line(color = "red")+ 
  geom_point(shape = 5) +
  scale_x_continuous(breaks = 2000:2025, minor_breaks= NULL)+ #(breaks = seq(2000, 2025, by =5), minor_breaks = NULL)
  theme(axis.text.x = element_text(angle = 60, hjust =1)) + 
  labs(title = "GDP per Capita in the USA, 2000-2025", 
       x = "year",
       y = "GDP")

#Histograms. shows distribution of continuous variables. Stores things like bins. should try histogram for the birds data with scale x. 
gap %>% 
  filter(year ==1935) %>% 
  ggplot(aes(x = lex)) + 
  geom_histogram(color = "black", fill="turquoise4", bins = 30) + 
  labs(title="Distribution of Life Expectancy")

#density plots. Uses smoothing function though like histogram to make distribution more clean looking 
gap %>% 
  filter(year ==1935) %>% 
  ggplot(aes(x = lex)) + 
  geom_density(color = "black", fill="turquoise4", alpha = 0.4) + 
  labs(title="Distribution of Life Expectancy")

#can put histogram and density plot together... 
gap %>% 
  filter(year ==1935) %>% 
  ggplot(aes(x =lex)) +
  geom_histogram(aes(y=after_stat(density)), color = "black", fill = "turquoise4") + 
  geom_density(color = "red", alpha = 0.4) 


#BIRD - could do something like this for most different months.... to show difference between seasons. 
gap %>% 
  filter(year==1930 | year == 1950) %>% #could do year %n% c(1930, 1950)
  mutate(year = factor(year)) %>% #they need to be in factors for them to be sepearte! 
  ggplot(aes(x = lex, color = year, fill=year)) + 
  geom_density(alpha = 0.3) + 
  labs(title ="Distribution of Life Expectancy, 1930 and 1950", 
       x = "Life Expectancy (years)",
       y = NULL,
       color = "Year",
       fill = "Year")

#creating categories with mutate and case_when... from dplyr.
income_cat2000 <- gap %>% 
  filter(year ==2000) %>%#creates new income categories based on data. 
  mutate(income_category = case_when(gdp<=1100 ~ "low",
                                     gdp > 1100 & gdp <= 4500 ~ "low_middle",
                                     gdp > 4500 & gdp <=13900 ~ "upper_middle",
                                     gdp > 13900 ~ "high")) %>% #looks at numeric and categorizes it
  select(country, gdp, income_category) %>% 
  mutate(income_category = factor(income_category,
                                  levels=c("low",
                                           "low_middle",
                                           "upper_middle",
                                           "high")))
```

## Plots and Aesthetics 

```{r}
#adding lines to graph (grid)
p + theme_linedraw()+ #draws in the lines 
  theme(axis.text.x = element_text(angle = 60, hjust=1))

#q
q <- homerange %>% 
  drop_na(log10_mass, log10_preymass) %>% #drop na first! 
  ggplot (aes(x=log10_mass, y=log10_preymass, color = thermoregulation)) + 
  geom_point (na.rm = T) + 
  geom_smooth (method = lm, se = F, color = "blue")
q
#making it clean
q + theme_classic()

#legneds. 
p+theme_linedraw()+
  theme(legend.position = "bottom", #top, bottom, left, right, none
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")

#ggthemes. library(ggthemes) has more!
p + 
  theme_few()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild") +
  theme_excel()

#some themes
theme_solarized
theme_foundation
theme_clean
theme_pander
theme_few
theme_stata
theme_par
#etc. etc. 

#RcolorBrewer 
#scale_colour_brewer is for points, scale_fill_brewer() is for fills 
p+scale_fill_brewer(palette = "Accent")+ #added brewer
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild") 

#Paleteer  - used to buid custom palette
my_palette <- paletteer_d("ggprism::plasma") #package::palette

#in action 
p+scale_fill_manual(values=my_palette)+ #calls the palette
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")

#can also use hex code to custom color... 
my_cols <- c("#66CDAA",
             "#FFCD00FF")
show_col(my_cols)

#faceting 
#AMAZING. allws you to make multi-panel plots for easy comparison 
homerange %>% 
  ggplot(aes(x = taxon, y = log10_mass, fill = thermoregulation)) +
  geom_boxplot() + 
  facet_wrap(~trophic_guild, ncol= 2, labeller = label_both) + #use nrow = 2 to make it up and down 
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position = "none",
        strip.text = element_text())
#facet_wrap makes a ribbon of panels by specified categorial variable. u can arrange them as wanted 
#facet_grid allows control over the faceted variables. can make comparison of two categorical variables. a~b, a is rows, b is columns 
homerange %>% 
  ggplot(aes(x = taxon, y = log10_mass, fill = thermoregulation)) +
  geom_boxplot() + 
  facet_grid(trophic_guild~thermoregulation, labeller = label_both) + #can do two things at once..
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position = "none",
        strip.text = element_text(size = 8, face ="bold"))
#BIRDS. facet maybe for cloud, etc. 

#random stuff. 
gapminder %>% 
  distinct(country) %>% #takes out distinct country
  n_distinct() #outputs the number 
```

##Dealing with NA's
```{r}
#first see if there are NA's
glimpse(life_history)
summary(life_history) #make sure values that shouldn't be negative arent 

life_history %>% 
  summarize(across(everything(), ~sum(is.na(.))))

#sometimes NA can be represented as -999 or even 0 
#replacing as NA... 
life_history_no_nas <- life_history_no_nas %>% 
  mutate(max_life=na_if(max_life, 0)) #check before updating 

#summary of NA across data frame
miss_var_summary(life_history_no_nas) #number and percent
#visual to help evaluate
life_history_no_nas %>% 
  gg_miss_var()

#or as a scatter plot... 
life_history_no_nas %>% 
  ggplot(aes(x=log10(mass), y = log10(wean_mass))) + 
  geom_point()+ 
  geom_miss_point() #idea of how much in each is missing

life_history_no_nas %>% 
  group_by(order) %>% 
  miss_var_summary(order =T)

#what to replace with na.... 
life_history %>% #going back to the original data
  replace_with_na(replace = list(newborn = "not measured", 
                                 weaning= -999, 
                                 wean_mass= -999, 
                                 afr= -999, 
                                 max_life= 0, 
                                 litter_size= -999, 
                                 gestation= -999, 
                                 mass= -999)) %>% 
miss_var_summary()

#or with a specific value 
life_history %>% 
  replace_with_na_all(condition = ~.x == -999) %>% 
  miss_var_summary()

#easiest way to do it lwk 
life_history %>%
  replace_with_na_all(condition = ~.x %in% c(common_na_numbers, common_na_strings)) %>%
  mutate(newborn = na_if(newborn, "not measured")) #to na if "not measured"
```

##Joining and formatting data in data frame
```{r}
#join_type(firstTable, secondTable, by=columnTojoinOn)

#inner join keeps rows that have matching values in both tables. ... if only exist in one = drop
inner_join(table_A, table_B, by = "customer_ID")

#left join keeps all row in first (left) table, and only matching rows in the second. if no match fills in with NA
left_join(table_A, table_B, by = "customer_ID")

#right join.... opposite of left join 
right_join(table_A, table_B, by = "customer_ID")

#full join keeps all rows in both tables and fills in NA where no match 
full_join(table_A, table_B, by = "customer_ID")

#anti join provides row in frst table for which there are not matching values in the second table 
anti_join(table_A, table_B, by = "customer_ID") #use to troubleshoot prolematic joint

```

## dates
```{r}
today() #current date
now() #current date and time 

#date and time can be formatted in multiple ways... 20200922 170000 etc. 
#can convert each using lubricate
mdy(09-22-2020)
dmy(22092007)
mdy_hms(month day year hour min sec)
#basically changes it to r format 

#saving final data to a '.csv'
write.csv(spider_date_cleaned,"spiders_with_locs.csv", row.names=FALSE)
```

##Maps
```{r}
#BBOX 
#figure out dimensions. 
spiders %>% 
  select(latitude,longitude) %>% 
  summary()
#create bbox 
lat <- c(34.67, 41.8)
long <- c(-124.1,-115.5)
bbox <- make_bbox(long, lat, f=.03)
#make map. 
map1 <- get_stadiamap(bbox, maptype = "stamen_terrain", zoom=7)
#taking a look 
ggmap(map1)
#adding points 
ggmap(map1)+
  geom_point(data=spiders, aes(longitude, latitude), size=.5, color="black", alpha=.8)+ #always long and then lat 
  labs(x="Longitude",y="Latitude",title="CA Usofila records")

#LEAFLET - interactive map 
spiders %>% 
  summarise(
    lon_min=min(longitude, na.rm = T),
    lon_max=max(longitude, na.rm = T),
    lat_min=min(latitude, na.rm = T),
    lat_max=max(latitude, na.rm = T))
#add to new variable
lon_min <- -124.13
lon_max <- -115.49
lat_min <- 34.67
lat_max <- 41.8
#build
leaflet(spiders) %>% 
  addProviderTiles(providers$Stadia.StamenTerrain) %>% 
  addCircleMarkers(
    lng = ~longitude, 
    lat = ~latitude,
    radius = 2,
    stroke = F,
    fillOpacity = .7
  ) %>% 
  addScaleBar(position="bottomleft") %>% 
  fitBounds(lng1 = lon_min, lat1 = lat_min,
            lng2 = lon_max, lat2 = lat_max)
```

##Shiny App 
```{r}
#skeleton 
#shortcut is 'shinyapp + shift + tab
ui <- fluidPage(
  
)

server <- function(input, output, session) {
  
}

shinyApp(ui, server)

#example done 

#UI 
ui <- fluidPage(
  selectInput("x",
              "Select X Variable",
              choices = c("bill_length_mm",
                          "bill_depth_mm",
                          "flipper_length_mm",
                          "body_mass_g"),
              selected="bill_length_mm"),
  
  selectInput("y",
              "Select Y Variable",
              choices = c("bill_length_mm",
                          "bill_depth_mm",
                          "flipper_length_mm",
                          "body_mass_g"),
              selected="bill_depth_mm"),
  
  checkboxInput("show_lm",
                "Show best-fit line(lm)",
                value=FALSE),
  
  plotOutput("plot",
             width="500px",
             height="400px")
)

#SERVER
server <- function(input, output, session) {

  output$plot <- renderPlot({
    
    ggplot(penguins,
           aes(x = .data[[input$x]],
               y = .data[[input$y]]))+ #safer way to explicitly tell shiny to treat it like data
      geom_point(na.rm=T)+
      if (input$show_lm)
        geom_smooth(method="lm", se=F, na.rm=T)
    else
      NULL
            })
}

#RUN APP 

shinyApp(ui, server)

#^ scatter plot. 
# for bar plot... 
ui <- fluidPage(
  
  radioButtons("x",
               "Select Fill Variable",
               choices=c("trophic_guild",
                         "thermoregulation"),
               selected="trophic_guild"),
  
  plotOutput("plot")
)

server <- function(input, output, session) {
  
  output$plot <- renderPlot({
    
    ggplot(data=homerange,
           aes_string(x="locomotion", fill=input$x))+
      geom_bar(position="dodge", alpha=0.8, color="black")+
      labs(x=NULL, y=NULL, fill="Fill Variable")
  })
  
}

shinyApp(ui, server)

#box plot wiht theme 
ui <- fluidPage(
  
  theme=shinythemes::shinytheme("cosmo"), #can add default themes. search up shiny theme. 
  
  selectInput("y",
              "Select Sleep Variable",
              choices = c("sleep_total", "sleep_rem", "sleep_cycle", "awake"),
              selected = "sleep_total"),
  
  plotOutput("plot", width = "600px", height = "500px")
)

server <- function(input, output, session) {
  output$plot <- renderPlot({
  msleep %>% 
    filter(vore != "NA") %>% 
      ggplot(aes(x = vore,
                 y = .data[[input$y]],
             fill = vore)) + 
      geom_boxplot(alpha = 0.75) +
      labs(title = "Sleep Variables by Vore Type",
           x = "Vore",
           fill = "Vore Type") +
      theme_minimal()
  })
}

shinyApp(ui, server)

#density 
ui <- fluidPage(
  titlePanel("Distribution by Order"),
  theme=shinythemes::shinytheme("cosmo"), #can add default themes. search up shiny theme. 
  
# helptext("asdfsdf") <- can use for source 
  
  selectInput("x",
              "Select Order",
              choices = sort(unique(gestation$order)), #gets rid of mistakes by sorting it and getting unique
              selected = "Afrosoricida"),
  
  selectInput("y",
              "Select Variable of Interest",
              choices = c("bodymass", "longevity", "gestationlength"),
              selected = "gestationlength"),
  
  plotOutput("plot", width = "600px", height = "500px")
)

server <- function(input, output, session) {
  output$plot <- renderPlot({
    
    gestation %>% 
      filter(input$x == .data[["order"]]) %>%
      ggplot(aes(.data[[input$y]])) + 
      geom_density(alpha = 0.4, fill = "green4", color = "black") + 
      theme_minimal() + 
      theme(plot.title = element_text(hjust =0.5))
  })
}

shinyApp(ui, server)

#side bar and other aesthetics 
ui <- dashboardPage(
  
  dashboardHeader(title = "Distribution of Gestation Length by Order"),
  
  dashboardSidebar(
    
     selectInput("x",
              "Select Order",
              choices = sort(unique(gestation$order)), #gets rid of mistakes by sorting it and getting unique
              selected = "Afrosoricida"),
    
    selectInput("y",
              "Select Variable of Interest",
              choices = c("bodymass", "longevity", "gestationlength"),
              selected = "gestationlength")
    ), #puts above into sidebar 
  
  dashboardBody(  
    plotOutput("plot", width = "600px", height = "500px")
  )
  
)

server <- function(input, output, session) {
  output$plot <- renderPlot({
    
    gestation %>% 
      filter(input$x == .data[["order"]]) %>% #basically sets the x variable columns 
      ggplot(aes(.data[[input$y]])) + 
      geom_density(alpha = 0.4, fill = "green4", color = "black") + 
      theme_minimal() + 
      theme(plot.title = element_text(hjust =0.5))
  })
}

shinyApp(ui, server)
```