---
title: "Homework 11"
author: "Jina Moon"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above. For any included plots, make sure they are clearly labeled. You are free to use any plot type that you feel best communicates the results of your analysis.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(ggthemes) #added for theme
``` 

```{r}
options(scipen = 999)
```

## Data
For this homework, we are going to use the [Gapminder](https://www.gapminder.org/data/) data on population (pop), GDP per capita (gdp), and life expectancy (lex). These data were downloaded and compiled using the `gapminder_compile.Rmd` script in the data folder.  

Let's focus our attention on data from the past 100 years, between 1925 and 2025.
```{r message=FALSE, warning=FALSE}
gapminder <- read_csv("data/gapminder.csv") %>% 
  clean_names() %>% 
  filter(between(year, 1925, 2025))
```

1. Use the function(s) of your choice to get an idea of the overall structure of the data, including its dimensions, column names, classes, etc. 
```{r}
glimpse(gapminder)
```

2. How many countries are represented in this dataset?
```{r}
gapminder %>% 
  distinct(country) %>% #takes out distinct country
  n_distinct() #outputs the number 
```
There are 195 countries represented in the dataset. 

3. Which five countries have had the largest absolute population growth between 1925-2025? Show this as a table.
```{r}
gapminder %>%
  filter(year == 1925 | year == 2025) %>% 
  select(country, year, pop) %>% 
  pivot_wider(names_from="year",
              values_from = "pop") %>% 
  mutate(pop_growth = `2025` - `1925`) %>% 
  arrange(desc(pop_growth)) %>% 
  head(n = 5)
```

India, China, Indonesia, Pakistan, and the USA had the largest absolute population growth between 1925-2025. 


4. Make a plot that shows population growth over time for the top country you found in question #3.
```{r}
#top country: India 
gapminder %>% 
  filter(country == "India") %>% 
  arrange(year) %>% 
  ggplot(aes(x= year, y = pop)) +
  geom_point(na.rm = T, size = 1) +
  geom_smooth(method = lm, se = F, color = "green4") + 
  labs(title = "Populatin Growth Over Time, India", x = "Year", y = "Population") + 
  theme(plot.title = element_text(hjust =0.5)) +
  theme_stata()

```

5. Has global life expectancy changed between 1925 and 2025? Show the min, median, mean, and max for all countries (combined) in the dataset.
```{r}
#for all combined
gapminder %>%
  filter(year == "1925" | year == "2025") %>% #looking at just these two so that it's not overwhelming
  group_by(year) %>%
  summarize(min = min(lex, na.rm = T),
            median = median(lex, na.rm = T),
            mean = mean(lex, na.rm = T),
            max = max(lex, na.rm = T))
```


Yes.... the global life expectancy has tremendously changed between 1925... People live a lot longer now (we basically have 30 more years to live life)

6. Make a plot that shows the distribution of life expectancy for all countries in 1925 and 2025 (try using `geom_density`). What do you notice about the shift in life expectancy over time?
```{r}
#two plots in the same graph
gapminder %>%  
  filter(year == 1925 | year == 2025) %>% 
  mutate(year = factor(year)) %>% 
  ggplot(aes(x = lex, color = year, fill = year))+ 
  geom_density(alpha = 0.4, na.rm = T, linewidth = 1) + 
  labs(title ="Distribution of Life Expectancy, 1925 and 2025", 
       x = "Life Expectancy (years)",
       y = NULL,
       color = "Year",
       fill = "Year") +
  theme(plot.title = element_text(hjust =0.5)) +
  theme_stata()

```

7. We are interested in the relationship between per capita GDP and life expectancy; i.e. does having more money help you live longer in 2025? Show this as a plot. (Suggestion! Remove Monaco as an extreme outlier.)
```{r}
gapminder %>% 
  filter(year == 2025 & country != 'Monaco') %>% #remove Monaco as extreme outlier 
  ggplot(aes(x = gdp, y = lex)) +
  geom_point (na.rm = T) + 
  geom_smooth (method = lm, se = F, color = "green4", na.rm = T) + 
  labs (title = "Per Capita GDP in Relation to Life Expectancy, 2025", x = "GDP", y = "Life Expectancy (years)") + 
  theme(plot.title=element_text(hjust = 0.5)) +
  theme_stata()
```

8. Which five countries have had the highest absolute GDP per capita growth over the past 50 years? Show this as a table. (Suggestion! Remove Monaco as an extreme outlier.)
```{r}
gdp_growth_gapminder <- gapminder %>% 
  filter((year == 2025 | year == 2025-50) & country != "Monaco") %>% #effectively 2025 and 1975, removed Monaco
  select(country, year, gdp) %>% 
  pivot_wider (names_from = "year",
               values_from="gdp") %>% #194 countries after taking out one so runs correctly 
  mutate(gdp_growth = `2025` - `1975`) %>% 
  arrange(desc(gdp_growth)) %>% 
  head(n = 5)
gdp_growth_gapminder
```

9. How does per capita GDP growth compare between these same five countries over the past 50 years? Show this as a plot.
```{r}
gdp_growth_gapminder %>% 
  ggplot(aes(x = reorder(country, - gdp_growth), y = gdp_growth)) + #reordered from greatest to smallest
  geom_col(aes (fill = country)) + 
  labs(title = "Per Capita GDP Growth in Top Five Countries Over the Past 50 Years", x = "Country", y = "Per Capita GDP Growth", fill = "Country") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme_stata()
```

10. Do one analysis of your choice that includes a table and plot as outputs.
```{r}
#Countries with most change in life expectancy in the last 50 years (top 10)

lex_growth_gapminder <- gapminder %>% 
  filter(year == 2025 | year == 2025-50) %>% #effectively 2025 and 1975
  select(country, year, lex) %>% 
  pivot_wider (names_from = "year",
               values_from="lex") %>% 
  mutate(lex_growth = `2025` - `1975`) %>% 
  arrange(desc(lex_growth)) %>% 
  head(n = 10)
lex_growth_gapminder
```

```{r}
lex_growth_gapminder %>% 
  ggplot(aes(x = reorder(country, - lex_growth), y = lex_growth)) + #reordered from greatest to smallest
  geom_col(aes (fill = country)) + 
  labs(title = "Life Expectancy Growth in Top 10 Countries Over the Past 50 Years", x = "Country", y = "Life Expectancy Growth", fill = "Country") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

## Knit and Upload
Please knit your work as an .html file and upload to Canvas. Homework is due before the start of the next lab. No late work is accepted. Make sure to use the formatting conventions of RMarkdown to make your report neat and clean! 