---
title: "Homework 4"
author: "Jina Moon"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
---

## Instructions
Answer the following questions and/or complete the exercises in RMarkdown. Please embed all of your code and push the final work to your repository. Your report should be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run. Any plots must have appropriate titles and aesthetics.  

## Load the tidyverse
```{r}
library("tidyverse")
options(scipen = 999) #disable scientific notation
```

For this homework, we will be using the `msleep` dataset from the `ggplot2` package. This dataset contains information about the sleep habits of various mammals. [Reference](https://ggplot2.tidyverse.org/reference/msleep.html): V. M. Savage and G. B. West. A quantitative, theoretical framework for understanding mammalian sleep. Proceedings of the National Academy of Sciences, 104 (3):1051-1056, 2007.

**1. Use a summary function of your choice to get an idea of the structure of the msleep data.**
```{r}
glimpse(msleep)
```
```{r}
msleep <- msleep #reads data into R to visualize 
```

**2. What are the names of the variables? To get a better idea of what the variables mean, you can use `?msleep`.**
```{r}
names(msleep)
?msleep
```

**3. Make a plot that shows the number of mammals in each vore type. Which vore type is most represented in the data?**
```{r}
names(msleep)
```
```{r}
#categorical and discrete .. so bar plot? 
ggplot(data=msleep, #specified data 
       mapping=aes(x=vore))+ #mapped aesthetics 
  geom_bar(mapping = aes(fill=vore))+ #added color to bar
  labs(title="Number of mammals in each vore type", x="vore type", y="count")
```
Herbivores are the most represented in our data. 

**4. What is the average length of time that each mammal sleeps?**
```{r}
mean(msleep$sleep_total)
```


**5. Use `summary()` to get a quick idea of the range of values for the variable `bodywt`.**
```{r}
summary(msleep)
```

**6. Build a plot that shows the range of `bodywt.` by vore type. Because the range of `bodywt` is so variable, you will need to scale the y-axis. You do this by adding `scale_y_log10()` as another layer.**
```{r}
#continuous and categorical.. so box plot
ggplot(data=msleep, mapping=aes(x=vore, y=bodywt))+
  geom_boxplot(mapping=aes(color=vore))+ #kind of unneccessary for this plot, yes, but still fun to do-- 
  scale_y_log10()+ #scale y-axis
  labs(title = "Body Weight (kg) by vore type", x="vore type", y="body weight (kg)")
```

**7. Is there a relationship between body weight and total sleep time? Add appropriate labels and a title to your plot.**
```{r}
names(msleep)
```

```{r}
#relationship ... both continous so scatter plot
ggplot(data=msleep, mapping=aes(x=bodywt, y=sleep_total))+
  geom_point()+
  geom_smooth(method=lm, se=T)+ #added line of best fit, with standard error
  scale_x_log10()+ #scaled the x-axis this time.. 
  labs(title="Body Weight (kg) vs. Total Sleep Time (hrs)", x="Body Weight (kg)", y="Total Sleep Time (hrs)")
```
Yes, there seems to be negative relationship between body weight and total sleep time... as body weight increases, the less sleep time overall 
**8. Which vore type tends to sleep the most? Build a plot to support your answer.**
```{r}
names(msleep)
```

```{r}
#comparing vore (categorical) to average sleep... so box plot 
ggplot(data=msleep, mapping = aes(x=vore, y=sleep_total))+ 
  geom_boxplot(mapping=aes(color=vore))+ 
  labs(title="Total sleep times by vore type", x= "vore type", y="total sleep time (hrs)")
```
Comparing the medain, the insectivore vore type seems like they sleep the most. 

**9. Is there a relationship between brain weight and body weight? Build a plot to support your answer.**
```{r}
#looking at relationship so scatter plot! 
ggplot(data=msleep, mapping=aes(x=brainwt, y=bodywt))+
  geom_point()+
  scale_x_log10()+
  scale_y_log10()+ #scaled both the y and x axis this time 
  geom_smooth(method=lm, se=T)+ #added line of best fit, with standard error
  labs(title="Brain Weight (kg) vs. Body Weight (kg)", x="Brain Weight (kg)", y="Body Weight (kg)")
```
Yes there seems to be a positive relationship between Brain weight and Body weight. 

**10. Build one plot of your choice that provides additional insight into the msleep dataset. Be sure to include appropriate labels and a title.**
```{r}

```

## Knit and Upload
Please knit your work as an .html file and upload to Canvas. Homework is due before the start of the next lab. No late work is accepted. Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  