---
title: "Homework 6"
author: "Jina Moon"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
---

## Instructions
Answer the following questions and/or complete the exercises in RMarkdown. Please embed all of your code and push the final work to your repository. Your report should be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run.  

## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Data 
For this assignment, we will use data from a study on vertebrate community composition and impacts from defaunation in [Gabon, Africa](https://en.wikipedia.org/wiki/Gabon). One thing to notice is that the data include 24 separate transects. Each transect represents a path through different forest management areas.  

Reference: Koerner SE, Poulsen JR, Blanchard EJ, Okouyi J, Clark CJ. Vertebrate community composition and diversity declines along a defaunation gradient radiating from rural villages in Gabon. _Journal of Applied Ecology_. 2016. A description of the variables is included inside the data folder.   

**1. Load `IvindoData_DryadVersion.csv` and store it as a new object called `gabon`.**
```{r}
gabon <- read.csv("data/ivindoData_DryadVersion.csv")
```

**2. Use one or more of the summary functions you have learned to get an idea of the structure of the data.**  
```{r}
glimpse(gabon)
```

**3. Clean the data as necessary following the conventions from lab 6.**
```{r}
gabon <- gabon %>% 
  clean_names()%>% #cleans col names
  mutate(across(where(is.character), tolower)) #cleans observations to lower case 
```

```{r}
glimpse(gabon) #make sure changes made
```

**4. Use `mutate()` Change the variables `HuntCat`, `LandUse`, and `TransectID` to factors. Be sure to store the output as a new dataframe and use it for the remaining questions.**
```{r}
names(gabon)
gabon <- gabon %>%
  mutate(across(c(hunt_cat, land_use, transect_id), as.factor))
glimpse(gabon)
```

**5. Make a plot that shows the number of transects for each land use type.**
```{r}
#count so bar plot
ggplot(data=gabon, 
       mapping=aes(x=land_use)) + 
  geom_bar(mapping=aes(fill=land_use)) + #bar plot
  labs(title = "Number of Transects for Each Land Use Type", x = "Land Use Type", y = "Count", fill= "Land Use Type")
```


**6. How does vegetation richness compare among land use types?**
```{r}
#continuous and categorical so box plot
ggplot(data = gabon, 
       mapping = aes(x=land_use , y = veg_rich)) +
  geom_boxplot(mapping = aes(fill = land_use)) + 
  labs(title = "Land Use Type vs. Vegetation Richness (no. tree species)", x = "Land Use Type", y = "Vegetation Richness (no. tree species)", fill = "Land Use Type")
```

Park tends to have more vegetation richness compared to logging or others, whereas logging has slightly more chance of having greater vegetation richness than others. 


**7. For which land use type (national parks, logging, or neither) is average all species diversity the greatest?**
```{r}
names(gabon) #idea of columns 
```

```{r}
gabon %>% 
  select(land_use, diversity_all_species) #view columns being worked with 
```

```{r}
#technically, not "average" since median, so I'll do further calculation, but also general trend should be depicted via box plot 
ggplot(data = gabon, 
       mapping = aes(x=land_use , y = diversity_all_species)) +
  geom_boxplot(mapping = aes(fill = land_use)) + 
  labs(title = "Land Use Type vs. All Species Diversity", x = "Land Use Type", y = "All Species Diversity", fill = "Land Use Type")
```

```{r}
#actual math part of this 
park <- gabon %>%
  select(land_use, diversity_all_species) %>%
  filter(land_use=="park")
avg_park <- mean(park$diversity_all_species)
avg_park

logging <- gabon %>%
  select(land_use, diversity_all_species) %>%
  filter(land_use=="logging")
avg_logging <- mean(logging$diversity_all_species)
avg_logging

neither <- gabon %>%
  select(land_use, diversity_all_species) %>%
  filter(land_use=="neither")
avg_neither <- mean(neither$diversity_all_species)
avg_neither
```

The national park has the greatest average all species diversity at 2.425, and as reflected in the box plot in comparison to other land use. 

**8. Is there a relationship between vegetation richness and mammal species richness?**
```{r}
#scatter plot 
names(gabon) #for reference

ggplot(data = gabon, 
	mapping = aes(x=veg_rich, y=rich_mammal_species)) + 
	geom_point(na.rm=T) + 
	geom_smooth(method=lm, se=F) + 
  labs(title= "Vegetation Richness (no. tree species) vs. Mammal Species Richness (no. species)", x= "Vegetation Richness (no. tree species)", y = "Mammal Species Richness (no. species)") 
	
```

Yes, there seems to be a positive relationship between vegetation richness and mammal species richness.


**9. Use `filter` to find all transects that have greater than 15 tree species or a breast height diameter between 50 and 60cm.  **
```{r}
names(gabon)

gabon %>% 
  select(transect_id, veg_rich, veg_dbh) %>% #for easier viewing of needed columns 
  filter((veg_rich > 15) | between(veg_dbh, 50, 60))
```


**10.Which transects and land use types have more than 10 tree species and 10 mammal species? Use `arrange()` to sort by the number of tree species.**
```{r}
gabon %>% 
  select(transect_id, land_use, veg_rich, rich_mammal_species) %>% 
  filter(veg_rich > 10, rich_mammal_species > 10) %>% 
  arrange(veg_rich)
```

## Knit and Upload
Please knit your work as an .html file and upload to Canvas. Homework is due before the start of the next lab. No late work is accepted. Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  